\Large Clustering Vehicles based on Trips Identified from Automatic \\ Number Plate Recognition Camera Scans

### 2018-02-23 classification.tex
A few changes - submitted to supervisor for review

% !TEX root = paper.tex
\subsection{Clustering trips}\label{s.classification}

@@

One direct application of trip data is unsupervised learning. In this section we identify groups of vehicles with similar trip patterns based on frequency and diversity of travel. Clusters can represent private vehicles, such as work-home commuters, transit vehicles like buses and taxis, or other types of vehicles such as delivery trucks.
%This in turn, may enable trip modes to be estimated
%However, due to its limited coverage of the road network, it is not clear how well ANPR data can capture much the travel characteristics necessary to accurately describe the different groups.

Due to distinctly different traffic behaviour during weekends, we chose to consider only trips occurring during weekdays. Therefore, we used all number plate data collected between the 6th and 24th of February and excluded data from the 2 weekends in-between. Furthermore, as mentioned in section~\ref{s.trips}, we used fixed empirical values for $\Tau$. Table~\ref{t:trips-tau} displays the number of trips, average trip degree and the proportion of trips of degree one, for varying values of $\Tau$. The effect of incrementing $\Tau$ on resulting trips is clear: increased trip degrees and decreased trips containing just a single sighting. On the other hand, we did not set a value for $\tau$, but instead handled implausible trips by filtering all sightings with confidence below 85\%. Duplicates were filtered after identifying consecutive sightings by the same camera occurring within the same trip. Clock synchronisation errors were provided in milliseconds with none exceeding 5 seconds. These were hence ignored.

@@

Table~\ref{t:tau_comparison} provides a summary of multiple runs of \emph{k-means}. For each value of $\Tau$, the optimal number of clusters is selected by picking the value of $k$ that maximises the Calinski-Harabasz criterion. Furthermore, the measures of inter-cluster (betweeness) and intra-cluster (withiness) are depicted relative to the corresponding total sum of squares (total betweeness and total withiness). It is noteworthy that the best value of $k$ increases inversely to $\Tau$. Although a higher value of $k$ seems to suggest that trips with smaller values of $\Tau$ are able capture the variance in the data better, we have to consider that varying $\Tau$ affects the average trip degree in the same direction whilst affecting the total number of trips in the opposite direction (table~\ref{t:trips-tau}).

Tables~\ref{t:kmeans_centers_450} and~\ref{t:kmeans_centers_1200} depict the cluster centres for $\Tau$ equal to 7.5 minutes and 20 minutes respectively. These partially meet our expectations. For instance, we were expecting to find a relatively small cluster representing taxis with a high average number of trips per day, occurring over a variety of origins and destinations and over a large time frame. Clusters 2 and 7 for $\Tau = 7.5$ and cluster 4 for $\Tau = 20$ do indeed fit this profile. What differentiates cluster 2 from 7 in the first case is essentially a lower mean number of trips per day and smaller average time at which the first and last trips of the day occur. Additionally, buses to some extent fit this category as well, however we expected that buses could be separated from taxis by showing less diversity in the number of origins and destinations as these essentially do multiple runs of the same trip throughout the day. Still, this can be explained by the fact that buses take routes through main and secondary roads. As most cameras are placed in main roads, the one long bus trip can be perceived as multiple small trips as the bus alternates between arterial and main roads. This is in fact, one of the downsides of ANPR data and is something that the methodology presented in section~\ref{s.trips} needs to develop in the future.

On the other hand, we expected to observe a group representing home to work commuters with the first trip of the day starting approximately at eight in the morning and a second trip terminating between five and six in the evening. Although we observe one or two groups with those characteristics, these contain a higher average of trips per day than expected, which may represent for example work to school trips. However, we observe at least one big group of trips occurring mostly during lunch hour. Several interpretations are possible but further work is needed to provide a more consistent interpretation and validation of these results. A big contributing factor however is the fact that a high proportion of trips contains only a single sighting (table~\ref{t:trips-tau}). It may be the case that commuters choose other routes than those going through ANPR cameras. Still, we need to devise further models and methods that are able to capture this uncertainty.

@@

\begin{table}[t]
\centering
\tabcolsep=0.15cm
\begin{tabular}{c c c c c c}
  \hline
 Cluster & Size & \thead{Average\\Trips} & \thead{Average\\Distinct\\Origins} & \thead{Average\\First\\Hour} & \thead{Average\\Last\\Hour} \\
  \hline
1 & 157962 & 2.45 & 2.27 & 11.38 & 14.93 \\
  2 & 303513 & 2.18 & 2.09 & 12.53 & 14.47 \\
  3 & 21549 & 6.02 & 5.00 & 8.68 & 17.23 \\
  4 & 108094 & 2.99 & 2.73 & 9.99 & 15.86 \\
  5 & 509 & 33.62 & 10.42 & 5.81 & 19.67 \\
  6 & 1993 & 17.99 & 11.24 & 6.45 & 19.09 \\
  7 & 4971 & 10.41 & 7.64 & 7.97 & 18.04 \\
  8 & 58059 & 4.09 & 3.62 & 9.22 & 16.60 \\
   \hline
\end{tabular}
\caption{Cluster sizes and centres for $\Tau = 5$ minutes.}
\label{t:kmeans_centers_300}
\end{table}

>>>>

% !TEX root = paper.tex
\subsection{Clustering vehicles}\label{s.classification}

@@

One application of trip data is to group vehicles together according to similar trip patterns, namely frequency and diversity of travel. We can distinguish vehicles based on private and transit modes of transportation and classes within these, such as everyday commuters versus sporadic drivers, buses versus taxis, and other types of vehicles such as lorries and delivery trucks. Ideally, we would like to build a model, through supervised learning, that is able to classify a vehicle from unseen trip data. However, due to privacy laws, there is no publicly available database that maps a plate number into one or several of the categories above. Therefore, we've decided to group vehicles by performing unsupervised learning using the \emph{k-means} clustering algorithm. Despite not being able to perform a rigorous validation or interpretation of the resulting clusters, we can begin to understand how useful trip data is to describe vehicles' travel patterns, given the limited coverage of the cameras in the road network.

Due to distinctly different traffic behaviour during weekends, we only consider trips occurring during weekdays. Therefore, we used all number plate data collected between the 6th and 24th of February and excluded data from the 2 weekends in-between. Furthermore, as mentioned in section~\ref{s.trips}, we used fixed empirical values for $\Tau$. Table~\ref{t:trips-tau} displays the number of trips, average trip degree and the proportion of trips of degree one, for varying values of $\Tau$. The effect of incrementing $\Tau$ on resulting trips is clear: increased trip degrees and decreased trips containing just a single sighting. On the other hand, we did not set a value for $\tau$, but instead handled implausible trips by filtering all sightings with confidence below 85\%. Duplicates were filtered after identifying consecutive sightings by the same camera occurring within the same trip. Clock synchronisation errors were provided in milliseconds with none exceeding 5 seconds. These were hence ignored.

@@

Table~\ref{t:tau_comparison} provides a summary of multiple runs of \emph{k-means}. For each value of $\Tau$, the optimal number of clusters is selected by picking the value of $k$ that maximises the Calinski-Harabasz criterion. Furthermore, the measures of inter-cluster (betweeness) and intra-cluster (withiness) are depicted relative to the corresponding total sum of squares (total betweeness and total withiness). It is noteworthy that the best value of $k$ increases inversely to $\Tau$. Although a higher value of $k$ seems to suggest that trips with smaller values of $\Tau$ are better able to capture the variance of trip data, we have to consider that varying $\Tau$ affects the average trip degree in the same direction whilst affecting the total number of trips in the opposite direction (table~\ref{t:trips-tau}).

%This in turn, may enable trip modes to be estimated
%However, due to its limited coverage of the road network, it is not clear how well ANPR data can capture much the travel characteristics necessary to accurately describe the different groups.

Table~\ref{t:kmeans_centers_300} depicts the cluster centres for the combination of $\Tau$ and $k$ that maximises the Calinski-Harabasz criterion. These results partially meet our expectations. For instance, we were expecting to find a relatively small cluster representing taxis with a high average number of trips per day, occurring over a variety of origins and destinations and over a large time frame. Clusters 5, 6 and 7 do indeed fit this profile. What differentiates cluster 5 from 6 is the particularly high number of average trips per day. When $k$ is equal to 7 these two clusters merge into one. To some extent, other categories fit this profile as well, namely buses, lorries and delivery trucks. However, we could expect buses to show less diversity in the number of origins and destination as these essentially do multiple runs of the same route throughout the day. Still, this can be explained by the fact that buses take routes through main and secondary roads. As most cameras are placed in main roads, the one long bus trip can be perceived as multiple small trips as the bus alternates between arterial and main roads. This is in fact, one of the downsides of ANPR data and the methodology presented in section~\ref{s.trips}.

On the other hand, we expected to observe a group representing home to work commuters with the first trip of the day starting approximately at eight in the morning and a second trip terminating between five and six in the evening. Although we observe one or two groups with those characteristics, these contain a higher average of trips per day than expected, which may represent for example work to school trips. However, we observe at least one big group of trips occurring mostly during lunch hour. A big contributing factor however is the fact that a high proportion of trips contains only a single sighting (table~\ref{t:trips-tau}). It may be the case that many of these drivers choose routes other than those going through ANPR cameras.

To improve the interpretability of these results, we've gain access to a database that outputs the vehicle's weight, wheelplane, make and category from the corresponding plate number. Broadly speaking, vehicle categories are defined relative to the transport of persons or goods, and sub-categories are given according to maximum allowed mass. With this information we sorted vehicles in the ANPR database into one of the following classes: \textit{Bus, Car, Motorcycle, Truck, Van}. Unfortunately, we were unable to identify taxis and therefore were unable to separate private cars from taxis. Table~\ref{t:} shows the mean relative frequency of these classes, and the corresponding deviation for the clusters shown in table~\ref{t:kmeans_centers_300}. Only the clusters showing more notable deviations from the mean are shown. Cluster 5, in particular is mainly composed by trucks. Cluster 6, sees an increase of frequency of cars which may represent taxis. Vans are seen in larger proportions in Cluster 7 and 3. These results provide some evidence in favor of some of our assumptions above and demonstrate that whilst ANPR data is able to capture some of the travel patterns, a clear and robust assessment of travel patterns is still not possible. Furthermore, it is not clear whether this is due to limited coverage of the road network or to a lack of methods that model, quantify and address the uncertainty of trips identified from ANPR data.

%Several interpretations are possible but further work is needed to provide a more consistent interpretation and validation of these results.  The lack of baselines and validation data is hence one of the shortcomings of this study. An additional pitfall of this methodology is that it is unable to evaluate the reliability of extracted trips.

%Nevertheless, trip data derived from ANPR number plate data is able to capture . This opens up room for the development of further methods that model and quantify this uncertainty.

@@

\begin{table}[t]
\centering
\tabcolsep=0.15cm
\begin{tabular}{c c c c c c}
  \hline
 Cluster & Size & \thead{Average\\Trips} & \thead{Average\\Distinct\\Origins} & \thead{Average\\First\\Hour} & \thead{Average\\Last\\Hour} \\
  \hline
1 & 157,962 & 2.45 & 2.27 & 11.38 & 14.93 \\
  2 & 303,513 & 2.18 & 2.09 & 12.53 & 14.47 \\
  3 & 21,549 & 6.02 & 5.00 & 8.68 & 17.23 \\
  4 & 108,094 & 2.99 & 2.73 & 9.99 & 15.86 \\
  5 & 509 & 33.62 & 10.42 & 5.81 & 19.67 \\
  6 & 1,993 & 17.99 & 11.24 & 6.45 & 19.09 \\
  7 & 4,971 & 10.41 & 7.64 & 7.97 & 18.04 \\
  8 & 58,059 & 4.09 & 3.62 & 9.22 & 16.60 \\
   \hline
\end{tabular}
\caption{Cluster sizes and centres for $\Tau = 5$ minutes.}
\label{t:kmeans_centers_300}
\end{table}

### 2018-02-23 discussion.tex
A few changes - submitted to supervisor for review

Once trip data has been computed, a range of interesting applications are available. This work tries to identify groups of vehicles by clustering information about frequency and diversity of travel. By associating a vehicle with a cluster that represents taxis, or home-to-work commuters, one can begin estimating trip mode usage across the city. However, the results presented here could benefit from extra work and further validation. Finally, future work can focus on using trip data to solve interesting research problems such as:
\begin{enumerate*}[label=(\roman*)]
  \item real-time route recommendation using probabilistic graphical models;
  \item detection of abnormal trip patterns for helping law enforcement in the identification of suspect vehicles or behaviour;
  \item modelling how drivers make routing choices in the presence of anomalies in the road network.
\end{enumerate*}

>>>>

Once trip data has been computed, a range of interesting applications are available. This work tries to identify groups of vehicles by clustering information about frequency and diversity of travel. By associating a vehicle with a cluster that represents taxis, or home-to-work commuters, one can begin estimating trip mode usage across the city. However, the results presented here could benefit from extra work and further validation. Namely, gaining access to a database that maps plate numbers to vehicle types would not only provide means to better validate the proposed methodology but also enable the application of supervised learning instead. Furthermore, this work can be improved upon by using other data sources as a baseline for evaluating the reliability of extracted trips, such as GPS taxi traces. Finally, future work can focus on using trip data to solve interesting research problems such as:
\begin{enumerate*}[label=(\roman*)]
  \item real-time route recommendation using probabilistic graphical models;
  \item detection of abnormal trip patterns for helping law enforcement in the identification of suspect vehicles or behaviour;
  \item modelling how drivers make routing choices in the presence of anomalies in the road network.
\end{enumerate*}

\subsection*{Acknowledgments}

The authors would like to thank Phil Blythe, Professor of Intelligent Transport Systems at Newcastle University, and Ray King from the Tyne and Wear Urban Traffic and Management Centre for providing guidance and access to the Automatic Number Plate Recognition database.

### 2018-02-23 paper.tex
A few corrections and changes. Changed the name of length to degree, removed most references to future work. Rephrased a few sentences.
A few changes - submitted to supervisor for review

\title{\Large Clustering Trips Identified from Automatic Number \\ Plate Recognition Camera Scans}
\author{Pedro M. Pinto Silva \thanks{School of Computing, Newcastle University, United Kingdom}
\and
Matthew Forshaw\footnotemark[1]
\and
A. Stephen McGough\footnotemark[1]}
\date{}

@@

\begin{abstract} \small\baselineskip=9pt In major cities, government agencies increasingly employ automatic number-plate recognition (ANPR) technology in law enforcement and traffic control. In the Tyne and Wear region (UK) the network of ANPR cameras is used to monitor travel times across sensitive roads. However, few works have explored the full potential of number-plate scans for analysing individual and collective travel patterns. In this work we present a methodology for deriving trips from vehicle sightings at fixed camera locations. We illustrate the effect of parameters $\tau$ and $\Tau$ on trip discrimination and on the detection of implausible trips. To demonstrate the potential of trip data we apply \emph{k-means} clustering to trips identified from over 40 million plate scans recorded over fifteen weekdays. Results show that whilst private and transit travel modes can begin to be inferred from the resulting clusters, further work needs to be put into developing a more consistent and integrated framework for trip identification in ANPR data.\end{abstract}

>>>>

\title{\Large Clustering Vehicles based on Trips Identified from Automatic \\ Number Plate Recognition Camera Scans}
\author{Pedro M. Pinto Silva \thanks{School of Computing, Newcastle University, United Kingdom}
\and
Matthew Forshaw\footnotemark[1]
\and
A. Stephen McGough\footnotemark[1]}
\date{}

@@

\begin{abstract} \small\baselineskip=9pt In major cities, government agencies increasingly employ automatic number-plate recognition (ANPR) technology in law enforcement and traffic control. In the Tyne and Wear region (UK) the network of ANPR cameras is used to monitor travel times across sensitive roads. So far, few works have explored the full potential of number-plate scans for analysing individual and collective travel patterns. In this work we present a methodology for deriving trips from vehicle sightings at fixed camera locations. We illustrate the effect of parameters $\tau$ and $\Tau$ on trip discrimination and on the detection of implausible trips. To demonstrate the potential of trip data we apply \emph{k-means} clustering to trips identified from over 40 million plate scans recorded over fifteen weekdays. Results show that whilst private and transit travel modes can begin to be inferred from the resulting clusters, further work needs to be put into developing a more consistent and integrated framework for trip identification in ANPR data.\end{abstract}
